---
title: "Group 8 project"
author: "Group 8"
format:
  pdf:
    fontsize: 11pt
    documentclass: article
    geometry: margin=0.25in
    number-sections: true
editor: visual
---

```{r}
#| warning: false
#| echo: false
library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(kableExtra)
library(gridExtra)

```

# Introduction {#sec-intro}

This investigation uses data from the 2008 - 2012 Scottish Health Survey, which collects information on the health, socio-economic, and lifestyle factors of people living in Scotland. The factors included in this dataset are: age group, sex, employment status, meets recommended vegetable intake (Yes/No), meets recommended fruit intake (Yes/No), survey year, and body mass index (BMI).

Body mass index is a quantity derived from the mass and height of a person, calculated as mass divided by the squared height, and measured in $kg/m^2$. It is a useful indicator for determining if an individual is underweight (BMI under 18.5), overweight (BMI over 25), or at a healthy weight, and it is especially useful for health data.

The aim of this investigation is to determine whether the body mass index in Scotland has changed between 2008 and 2012, and to assess if there are any differences in the BMI distribution by age, gender, socio-economic factors (employment status) or lifestyle factors (meets recommended fruit and veg intake).

# Exploratory Analysis {#sec-eda}

```{r}
#| echo: false
BMI_data<-read.csv("DAProject8.csv")

```

## Data Summaries

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
BMI_F<-BMI_data%>%
  filter(Sex=="Female")
BMI_M<-BMI_data%>%
  filter(Sex=="Male")

any(is.na(BMI_data))
```

We first inspect the balance of the data and check for missing values. The dataset contains a total of 25224 participants. There were no missing values in any of the variables used in this report, so all 25224 participants were retained for both the exploratory analysis and the subsequent modelling.

@tbl-BMI_sum displays the summary statistics for BMI. The mean and median are very similar (27.81 and 27.17 respectively), indicating that the distribution is approximately symmetric, while the standard deviation of 5.26 suggests a large variation in BMI across individuals. The maximum and minimum values of 12.59 and 59.11 respectively indicate a wide range of BMI values within the sample.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-BMI_sum
#| tbl-cap: Summary statistics for BMI

summary_BMI<-BMI_data%>%
  summarise(Mean=round(mean(BMI_data$BMI),3),
             SD=round(sd(BMI_data$BMI),3),
             Median=round(median(BMI_data$BMI),3),
             Min=round(min(BMI_data$BMI),3),
             Max=round(max(BMI_data$BMI),3))

kable(summary_BMI)
```

## Distribution Of BMI

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-histogram
#| fig-cap: Histogram showing distribution of BMI.
ggplot(BMI_data,aes(x=BMI))+
  geom_histogram(binwidth=1,fill="hotpink1",color="black")+
  labs(x="BMI",
       y="Freguency")
```

@fig-histogram shows BMI's distribution is slightly right-skewed and peaks around 26-28. We can observe that the majority of values fall between 22-32, with a large portion of individuals being classifiable as overweight. Highlighted by the peak falling within the overweight grouping.

## BMI and Year

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6.5
#| fig-height: 2.5
#| fig-pos: 'h'
#| label: fig-boxplots_BMI_year
#| fig-cap: Boxplots of BMI for each year of survey
ggplot(BMI_data,aes(x=factor(Year),y=BMI,fill=factor(Year)))+
  geom_boxplot()+
  labs(x="Year", y="BMI (kg/m^2)")+
  theme(legend.position = "none")+
  scale_fill_manual(values=c("darkolivegreen3","hotpink1","mediumpurple2","gold","steelblue2"))
```

@fig-boxplots_BMI_year shows the distribution of BMI across the different years of the Scottish Health Survey. The median BMI appears to remain stable over the entire period with no clear trend. The interquartile ranges and overall spread of the data are similar across the entire period. There is little evidence to suggest that BMI changed over the study period, but formal analysis is required to confirm this.

## BMI and Additional Explanatory Variables

### Demographic Variables

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 2
#| fig-pos: 'h'
#| label: fig-boxplots_demo_vars
#| fig-cap: Boxplots of BMI across the demographic variables, age group and sex
p1<-ggplot(BMI_data,aes(x=AgeGroup,y=BMI,fill=AgeGroup))+
  geom_boxplot()+
  scale_fill_manual(values=c("darkolivegreen3","darkorange","hotpink1","steelblue2","gold","mediumpurple2","cadetblue2"))+
  labs(title="BMI vs Age Group",
       x="Age Group", y="BMI")+
  theme(legend.position = "none")
p2<-ggplot(BMI_data,aes(x=Sex,y=BMI,fill=Sex))+
  geom_boxplot()+
  scale_fill_manual(values=c("darkolivegreen3","hotpink1"))+
  labs(title="BMI vs Sex",
       x="Sex", y="BMI")+
  theme(legend.position = "none")
grid.arrange(p1,p2,ncol=2)
```

@fig-boxplots_demo_vars shows the distribution of BMI by age group and sex. There appears to be a very slight increase in the median BMI as age group increases, suggesting that older individuals tend to have higher BMI values. But the interquartile ranges have a significant overlap so the differences between age groups may be small or insignificant and should further investigated in later formal analysis.

There appears to be little difference in the distribution of BMI for female and male participants, other than in the interquartile ranges, with a larger range in the female participants, indicating greater variability in BMI among females compared to males.

### Socio-Economic and Lifestyle Variables

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 3.5
#| fig-cap: Boxplots of BMI across the socio-economic variable, employment
#| label: fig-boxplot_employ
#| fig-pos: "H"

ggplot(BMI_data,aes(x=Employment,y=BMI,fill=Employment))+
  geom_boxplot()+
  scale_x_discrete(labels=c("Doing something else"="Else","In full-time education"="FT Edu","In paid employment, self-employed or on gov't training"="Employed","Looking after home/family"="Carer","Looking for/intending to look for paid work"="Searching","Perm unable to work"="Unable","Retired"="Retired"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(legend.position = "none")+
  scale_fill_manual(values=c("darkolivegreen3","darkorange","hotpink1","steelblue2","gold","mediumpurple2","cadetblue2"))+
  labs(y="BMI (kg/m^2)")

```

@fig-boxplot_employ shows the distribution of BMI across the different employment categories. Individuals in full-time education appear to have a lower BMI compared to other employment groups, but the interquartile ranges for all groups substantially overlap, suggesting that the differences in BMI across these groups are relatively small and significance should be investigated with formal analysis. There also appear to be outliers at both extremes of BMI across the categories.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 2
#| fig-pos: "H"
#| fig-cap: Boxplots of BMI across the lifestyle variables fruit and veg
#| label: fig-boxplots_fruit_veg

p2<-ggplot(BMI_data,aes(x=Veg,y=BMI,fill=Veg))+
  geom_boxplot()+
  theme(legend.position = "none")+
  scale_fill_manual(values=c("darkolivegreen3","hotpink1"))+
  labs(y="BMI (kg/m^2)",title="BMI vs Veg")
p1<-ggplot(BMI_data,aes(x=Fruit,y=BMI,fill=Fruit))+
  geom_boxplot()+
  theme(legend.position = "none")+
  scale_fill_manual(values=c("darkolivegreen3","hotpink1"))+
  labs(y="BMI (kg/m^2)",title="BMI vs Fruit")

grid.arrange(p1,p2,ncol=2)

```

@fig-boxplots_fruit_veg shows the distribution of BMI by fruit and vegetable consumption. The boxplots in both factors appear very similar with the medians, and interquartile ranges being at similar levels. This suggests that there is no clear evidence of a difference in BMI for participants who consume their recommemded fruit and veg and those who don't, but this should be investigated in formal analysis.

# Formal Data Analysis {#sec-fda}


```{=tex}
\begin{align*}
BMI ~=~& \alpha + \beta_{AgeGroup} \cdot \mathbb{I}_{AgeGroup}(x) + \beta_{Sex} \cdot \mathbb{I}_{Sex}(male) \\
       & + \beta_{Employment} \cdot \mathbb{I}_{Employment}(x) + \beta_{Veg} \cdot \mathbb{I}_{Veg}(Yes) \\
       & + \beta_{Fruit} \cdot \mathbb{I}_{Fruit}(Yes) + \beta_{Year} \cdot \mathbb{I}_{Year}(x)
\end{align*}
```
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: tbl-lm
#| tbl-cap: Estimates of the regression model coefficient

full_model <- lm(BMI ~ AgeGroup + Sex + Employment + Veg + Fruit + Year, data = BMI_data)
tab_model(full_model,show.ci = F)
```

Looking at @tbl-lm, we can see the reduced model is:

```{=tex}
\begin{align*}
BMI ~=~& \alpha + \beta_{Age[25\text{-}34]} \cdot \mathbb{I}_{Age}([25\text{-}34]) + \beta_{Age[35\text{-}44]} \cdot \mathbb{I}_{Age}([35\text{-}44]) \\
       & + \beta_{Age[45\text{-}54]} \cdot \mathbb{I}_{Age}([45\text{-}54]) + \beta_{Age[55\text{-}64]} \cdot \mathbb{I}_{Age}([55\text{-}64]) \\
       & + \beta_{Age[65\text{-}74]} \cdot \mathbb{I}_{Age}([65\text{-}74]) + \beta_{Age[75+]} \cdot \mathbb{I}_{Age}([75+]) \\
       & + \beta_{Emp[\text{FT Edu}]} \cdot \mathbb{I}_{Emp}([\text{FT Edu}]) + \beta_{Emp[\text{Employed}]} \cdot \mathbb{I}_{Emp}([\text{Employed}]) \\
       & + \beta_{Emp[\text{Unable}]} \cdot \mathbb{I}_{Emp}([\text{Unable}]) + \beta_{Emp[\text{Retired}]} \cdot \mathbb{I}_{Emp}([\text{Retired}]) \\
       & + \beta_{Veg} \cdot \mathbb{I}_{Veg}(Yes)
\end{align*}
```
With this reduced model, the fitted regression model is:

```{=tex}
\begin{align*}
BMI ~=~& -17.9 + 1.18 \cdot \mathbb{I}_{Age}([25\text{-}34]) + 2.5 \cdot \mathbb{I}_{Age}([35\text{-}44]) + 2.77 \cdot \mathbb{I}_{Age}([45\text{-}54]) \\
       & + 3.28 \cdot \mathbb{I}_{Age}([55\text{-}64]) + 3.19 \cdot \mathbb{I}_{Age}([65\text{-}74]) + 2.33 \cdot \mathbb{I}_{Age}([75+]) \\
       & - 0.87 \cdot \mathbb{I}_{Emp}([\text{FT Edu}]) + 0.67 \cdot \mathbb{I}_{Emp}([\text{Employed}]) \\
       & + 1.18 \cdot \mathbb{I}_{Emp}([\text{Unable to work}]) + 0.66 \cdot \mathbb{I}_{Emp}([\text{Retired}]) \\
       & - 0.42 \cdot \mathbb{I}_{Veg}(Yes)
\end{align*}
```
The baseline category for this model is individuals aged 16--24, not in full-time education and not meeting the recommended daily vegetable intake. Interpreting the coefficients relative to this baseline: BMI increases substantially with age, peaking in the 55--64 age group with a coefficient of 3.28, before declining slightly in the older age groups. Individuals unable to work have the highest BMI relative to the baseline, with a coefficient of 1.18, while those in full-time education have a lower BMI than the baseline, with a coefficient of -0.87. Employed and retired individuals show small positive associations with BMI of 0.67 and 0.66 respectively. Finally, individuals who meet the recommended daily vegetable intake have a BMI that is on average 0.42 units lower than those who do not.

```{r}
#| echo: false
#| fig-cap: Final model diagnostic plots
#| label: fig-resids
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

performance::check_model(full_model,check=c("qq","normality","homogeneity","linearity"))

```

The model assumptions were assessed using the diagnostic plots displayed in @fig-resids. The linearity plot shows residuals scattered randomly around zero, suggesting the linearity assumption is reasonably satisfied. The homogeneity of variance plot similarly shows no strong pattern in the spread of residuals across fitted values, indicating constant variance is approximately met. However, the Q-Q plot and density plot indicate that the residuals are not normally distributed, displaying a left skew. This suggests some caution should be applied when interpreting the results of this model, though given the large sample size, the parameter estimates remain approximately valid by the Central Limit Theorem.

# Conclusion {#sec-conc}
